---
title: "compareScenarios"
author: "Falk Benke"
date: "2024-04-26"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{compareScenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`piamPlotComparison` is a lightweight framework to visually compare the results of multiple runs of models, such as [REMIND](https://github.com/remindmodel/remind) or [EDGE-T](https://github.com/pik-piam/edgeTransport).

It is used in conjunction with a reporting library, such as [remind2](https://github.com/pik-piam/remind2) or [reporttransport](https://github.com/pik-piam/reporttransport). The reporting library includes Rmarkdown-files (Rmd-files) with R-code (e.g., for creating plots, creating tables) as well as descriptive text following some conventions. The framework offers functionality that reads model runs and produces either HTML- and PDF-documents using the Rmd-files via the core function `compareScenarios`.

Aside from HTML- and PDF-documents as output, `compareScenarios` also allows to obtain a copy of the Rmd-files needed to create these outputs (by setting the argument `outputFormat = "Rmd"`). Rendering the resulting Rmd-files to PDF or HTML yields the same documents as calls to `compareScenarios()` with `outputFormat = "PDF"` or `"HTML"`. The advantage of being able to access these Rmd-files is the possibility to change the plotting code without changing the code of the reporting library. Moreover, Rmd-files can be used interactively in RStudio.

# Minimal setup

...

# cs_main.Rmd

Internally, `compareScenarios()` calls `rmarkdown::render()` on certain Rmarkdown-files (Rmd-files). To be more precise, `rmarkdown::render()` is called on [cs_main.Rmd](https://github.com/pik-piam/piamPlotComparison/blob/master/inst/compareScenarios/cs_main.Rmd). This file loads the data from the mif-files, preprocesses the data, and includes the section-Rmd-files from the project library at the very end (and optionally further Rmd-files provided by the user, see YAML-parameter `userSectionPath`).

## YAML-header

The file `cs_main.Rmd` starts with a YAML header marked by `---`. This header declares some basic information of the report, like its title and the output format. Furthermore, it contains a list `params`, which parameterizes the report. Among others, such parameters are the paths to the mif-files and certain properties that are shared for all plots in the report. Each such parameter can be changed by a respective argument in the call of `compareScenarios()`, see section *YAML Parameters* in the documentation of the function.

## Loading

The mif-files are loaded using `quitte::read.quitte()`. This function names the global region `"World"` (not `"GLO"` as `magclass::read.report()`). For each file, we obtain one data-frame with columns `model, scenario, region, variable, unit, period, value`. During preprocessing the data-frames are concatenated into one quitte object named `data`. In the sections, `data` is used to provide the data necessary for the plots.

If provided, also config files are loaded.

## Preprocessing

* Scenarios are renamed if the user specifies new names (using a named vector for the argument `mifScen` of `compareScenarios()` or by setting the parameter `mifScenNames` in the Rmd-files) or if scenario names are duplicate.
* The columns `period` (years) and `region` are filtered according to the parameters `yearsScen`, `yearsHist`, and `reg`.
* `|+|, |++|, |+++|, ...` are removed from variable names. See also `piamutils::deletePlus`. An additional column `varplus` is added to the data frame, which retains the original name including `+`. This column is used by `mip::showAreaAndBarPlotsPlus()`.

If you need to execute any additional code as part of the general setup, put it into a file `inst/compareScenarios/preprocessing.R` in the project library. Custom code will be executed after the general code.

## Global Variables 

Global variables are created in `cs_main.Rmd` and are intended to be used in the plot functions of the section-Rmd-files.

The quitte-object (data frame) `data` with columns `model, scenario, region, variable, unit, period, value, varplus` provides all data that may be plotted.
It has mostly factor-columns. Thus, a vector of all available regions can be obtained by `levels(data$region)`.

Some arguments of the `mip::show...()` plot functions use the functionality provided by the base-R functions `options()` and `getOption()`. In `cs_main.Rmd` there are calls to `options()`, setting values of the global options `mip.mainReg` and `mip.yearsBarPlot`. The plotting functions may have an argument, e.g., `mainReg` with default value `getOption("mip.mainReg")`. Thus, this argument does not have to be stated for each call of the plot function (assuming the respective call of `options()` sets it to the right value).

If you want to set `mip.histRefModel`, please do so in the `preprocessing.R` file of your project library.

# Section-Rmd-files

The section-Rmd-files follow the naming pattern `cs_NN_XXXX.Rmd`, where `NN` is replaced by a two digit number and `XXXX` is replaced by a short name of the section. If the YAML-parameter `sections` is set to `"all"`, the default, all sections of this naming pattern are included by `cs_main.Rmd`. Alternatively, `sections` can be set to a vector of individual sections in the form of `"NN_XXXX"` to only render these sections.

The section-Rmd-files consist of section and sub-section titles, marked by `#`, `##`, `###`, ..., and R-code chunks which create plots, usually by calling one of the `show...()`-functions in the `mip` package.

## userSectionPath

The user can provide one or more additional Rmd-files that are appended after the sections provided by the package. Setting `sections` to `NULL` and `userSectionPath` to a character-vector of paths to Rmd-files creates a fully user-defined report.

## Special Sections

There is a first section (section number 0) and a last section in the files `cs_00_info.Rmd` and `cs_99_further_info.Rmd`, respectively. 

### 0 Info {-}

This section contains a file reference table with scenario names and the paths to the respective mif files. It also show if a scenario is renamed. Renaming can be forced by the user by using a named character vector as argument `mifScen` where the names indicate the new scenario name, e.g.,

```{r}
compareScenarios(
  projectLibrary = "myProjectLib",
  mifScen = c(newName1 = "path/to/scen1.mif", newName2 = "path/to/scen2.mif"),
  ...)
```

Or, if two scenarios have the same name, they are renamed automatically.

Furthermore, this section displays a description of each scenario. This requires the YAML-parameter `cfgScen` to be set (see also [99 Further Info]). The description can be set in the `scenario_config*.csv`-file via a column `description`.

### 99 Further Info {-}

If `compareScenarios()` is provided with paths to the `config.Rdata` files of the scenarios via the YAML-parameter `cfgScen`, an analysis of the configurations of the scenarios is shown.

```{r}
compareScenarios(
  projectLibrary = "myProjectLib",
  mifScen = c("path/to/scen1.mif", "path/to/scen2.mif"),
  cfgScen = c("path/to/scen1/config.RData", "path/to/scen2/config.RData"),
  ...)
```

# Content Functions

## Plot Functions

The section Rmd-files mostly consist of simple calls of a `show...()`-function from the the [mip](https://github.com/pik-piam/mip) package. These functions are designed to make the creation of re-occurring kinds of plots in `comapreSceanrios` simple. Alternatively, one can build custom plots using `ggplot2`.

The `mip`-package contains following default plot functions for `comapareScenarios`: 

* `showAreaAndBarPlots()`
* `showAreaAndBarPlotsPlus()`
* `showLinePlots()`
* `showLinePlotsWithTarget()`
* `showMultiLinePlots`
* `showMultiLinePlotsByVariable()`
* `showRegiLinePlots()`

All but `showLinePlotsWithTarget()` can use the global option `mip.mainReg`.
`showAreaAndBarPlots()` and `showAreaAndBarPlotsPlus()` can use `mip.yearsBarPlot`.
`showMultiLinePlotsByVariable()` uses `mip.histRefModel`.

See the respective function documentation for more information and example plots.

## Table Functions

For creating tables in cs, see the documentation of `piamPlotComparison::calcTimeSeriesStats()` and `piamPlotComparison::showStatsTable()` and -- for an example -- the section **Tables of Significant Periods** in [`remind2/inst/compareScenarios/cs2_01_summary.Rmd`](https://github.com/pik-piam/remind2/blob/master/inst/compareScenarios/cs_01_summary.Rmd).

